---
title: "Disease Burden Analysis"
format: pdf
---

# Initializing
```{r, warning=FALSE, message=FALSE} 
#| label: Loading packages

library(mosaic)   
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readxl)
library(writexl)
library(plotly)
library(dplyr)
library(readr)
library(here)
library(rvest)
library(purrr)
library(conflicted)
conflict_prefer("max", "mosaic")
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
conflict_prefer("summarize", "dplyr")
conflict_prefer("mean", "mosaic")
```

```{r}
#| label: Setting working directory

## Adjust!
knitr::opts_knit$set(root.dir = here::here())
```

# Loading Data
```{r, message=FALSE}
#| label: Loading in disease burden data frames

Cardiovascular_Disease_Mortality_Burden_path <- here("Data_Raw/Disease_Mortality_Burden", "Heart_Liver_Kidney_Disease_Death.xlsx")
Cardiovascular_Disease_Mortality_Burden <- read_xlsx(Cardiovascular_Disease_Mortality_Burden_path, sheet = 1)

Liver_Disease_Mortality_Burden_path <- here("Data_Raw/Disease_Mortality_Burden", "Heart_Liver_Kidney_Disease_Death.xlsx")
Liver_Disease_Mortality_Burden <- read_xlsx(Liver_Disease_Mortality_Burden_path, sheet = 1)

Kidney_Disease_Mortality_Burden_path <- here("Data_Raw/Disease_Mortality_Burden", "Heart_Liver_Kidney_Disease_Death.xlsx")
Kidney_Disease_Mortality_Burden <- read_xlsx(Kidney_Disease_Mortality_Burden_path, sheet = 1)
# Source:  https://www.cdc.gov/nchs/hus/contents2020-2021.htm#Table-SlctMort
```

```{r}
#| label: Loading in organ transplantation data frames

HRSA_National_Data_path <- here("Data_Raw/Organ_Transplant", "Organ_Donation_and_Transplantation_Data.xlsx")
HRSA_National_Data <- read_xlsx(HRSA_National_Data_path, sheet = 2)
# Source: https://data.hrsa.gov/data/download?data=organ#organ

HRSA_State_Data_path <- here("Data_Raw/Organ_Transplant", "Organ_Donation_and_Transplantation_Data.xlsx")
HRSA_State_Data <- read_xlsx(HRSA_State_Data_path, sheet = 3)
# Source: https://data.hrsa.gov/data/download?data=organ#organ

HRSA_Donor_Demographics_path <- here("Data_Raw/Organ_Transplant", "Organ_Donation_and_Transplantation_Data.xlsx")
HRSA_Donor_Demographics <- read_xlsx(HRSA_Donor_Demographics_path, sheet = 4)
# Source: https://data.hrsa.gov/data/download?data=organ#organ

HRSA_Candidate_Registration_path <- here("Data_Raw/Organ_Transplant", "Organ_Donation_and_Transplantation_Data.xlsx")
HRSA_Candidate_Registration <- read_xlsx(HRSA_Candidate_Registration_path, sheet = 7)
# Source: https://data.hrsa.gov/data/download?data=organ#organ

HRSA_Waitlist_Removal_path <- here("Data_Raw/Organ_Transplant", "Organ_Donation_and_Transplantation_Data.xlsx")
HRSA_Waitlist_Removal <- read_xlsx(HRSA_Waitlist_Removal_path, sheet = 9)
# Source: https://data.hrsa.gov/data/download?data=organ#organ

HRSA_Recipient_Demographics_path <- here("Data_Raw/Organ_Transplant", "Organ_Donation_and_Transplantation_Data.xlsx")
HRSA_Recipient_Demographics <- read_xlsx(HRSA_Recipient_Demographics_path, sheet = 10)
# Source: https://data.hrsa.gov/data/download?data=organ#organ
```

# Cleaning Data

## Disease mortality burden by organ v sex and race

```{r, warning=FALSE}
#| label: Cleaning cardiovascular disease mortality burden data and creating by race/sex data frames

Sexes <- c("Male", "Female")
Races <- c("White only", "Black or African American only", "Asian or Pacific Islander only", "Hispanic or Latino")

# Renaming columns
Cardiovascular_Disease_Mortality_Burden_Columns <- as.character(Cardiovascular_Disease_Mortality_Burden[3, ])
Cardiovascular_Disease_Mortality_Burden_Columns[is.na(Cardiovascular_Disease_Mortality_Burden_Columns)] <- "NA"
Cardiovascular_Disease_Mortality_Burden_Columns <- make.unique(Cardiovascular_Disease_Mortality_Burden_Columns)
  # make.unique appends a sequence of numbers to make the column names unique
colnames(Cardiovascular_Disease_Mortality_Burden) <- Cardiovascular_Disease_Mortality_Burden_Columns

Cardiovascular_Disease_Mortality_Burden <- Cardiovascular_Disease_Mortality_Burden |>
  rename(Characteristic = "Sex and cause of death\\1") |>
  select(Characteristic,`1997\\3`:`2019\\4`)

# Removing irrelevant data
Cardiovascular_Disease_Mortality_Burden <- Cardiovascular_Disease_Mortality_Burden[c(6, 26, 47, 89, 129, 209, 269), ]

# Updating characteristics for "Chronic liver disease and cirrhosis"
Cardiovascular_Disease_Mortality_Burden[1, "Characteristic"] <- "All"
Cardiovascular_Disease_Mortality_Burden[2, "Characteristic"] <- "Male"
Cardiovascular_Disease_Mortality_Burden[3, "Characteristic"] <- "Female"
Cardiovascular_Disease_Mortality_Burden[4, "Characteristic"] <- "White only"
Cardiovascular_Disease_Mortality_Burden[5, "Characteristic"] <- "Black or African American only"
Cardiovascular_Disease_Mortality_Burden[6, "Characteristic"] <- "Asian or Pacific Islander only"
Cardiovascular_Disease_Mortality_Burden[7, "Characteristic"] <- "Hispanic or Latino"

# Cleaning year names
colnames(Cardiovascular_Disease_Mortality_Burden) <- gsub("\\\\[0-9]$", "", colnames(Cardiovascular_Disease_Mortality_Burden))

Cardiovascular_Disease_Mortality_Burden <- Cardiovascular_Disease_Mortality_Burden |>
  mutate(across(`1997`:`2019`, ~ as.numeric(.x) / 100000))

print(Cardiovascular_Disease_Mortality_Burden)

Cardiovascular_Disease_Mortality_Burden_Sex <- Cardiovascular_Disease_Mortality_Burden |>
  filter(Characteristic %in% c(Sexes)) |>
rename(Sex = Characteristic)

print(Cardiovascular_Disease_Mortality_Burden_Sex)

Cardiovascular_Disease_Mortality_Burden_Race <- Cardiovascular_Disease_Mortality_Burden |>
  filter(Characteristic %in% c(Races)) |>
  rename(Race = Characteristic)

print(Cardiovascular_Disease_Mortality_Burden_Race)
```

```{r, warning=FALSE}
#| label: Cleaning liver disease mortality burden data and creating by race/sex data frames

# Renaming columns
Liver_Disease_Mortality_Burden_Columns <- as.character(Liver_Disease_Mortality_Burden[3, ])
Liver_Disease_Mortality_Burden_Columns[is.na(Liver_Disease_Mortality_Burden_Columns)] <- "NA"
Liver_Disease_Mortality_Burden_Columns <- make.unique(Liver_Disease_Mortality_Burden_Columns)
  # make.unique appends a sequence of numbers to make the column names unique
colnames(Liver_Disease_Mortality_Burden) <- Liver_Disease_Mortality_Burden_Columns

Liver_Disease_Mortality_Burden <- Liver_Disease_Mortality_Burden |>
  rename(Characteristic = "Sex and cause of death\\1") |>
  select(Characteristic,`1997\\3`:`2019\\4`)

# Removing irrelevant data
Liver_Disease_Mortality_Burden <- Liver_Disease_Mortality_Burden[c(14, 35, 56, 97, 137, 217, 277), ]

# Updating characteristics for "Chronic liver disease and cirrhosis"
Liver_Disease_Mortality_Burden[1, "Characteristic"] <- "All"
Liver_Disease_Mortality_Burden[2, "Characteristic"] <- "Male"
Liver_Disease_Mortality_Burden[3, "Characteristic"] <- "Female"
Liver_Disease_Mortality_Burden[4, "Characteristic"] <- "White only"
Liver_Disease_Mortality_Burden[5, "Characteristic"] <- "Black or African American only"
Liver_Disease_Mortality_Burden[6, "Characteristic"] <- "Asian or Pacific Islander only"
Liver_Disease_Mortality_Burden[7, "Characteristic"] <- "Hispanic or Latino"

# Cleaning year names
colnames(Liver_Disease_Mortality_Burden) <- gsub("\\\\[0-9]$", "", colnames(Liver_Disease_Mortality_Burden))

Liver_Disease_Mortality_Burden <- Liver_Disease_Mortality_Burden |>
  mutate(across(`1997`:`2019`, ~ as.numeric(.x) / 100000))

print(Liver_Disease_Mortality_Burden)

Liver_Disease_Mortality_Burden_Sex <- Liver_Disease_Mortality_Burden |>
  filter(Characteristic %in% c(Sexes))

print(Liver_Disease_Mortality_Burden_Sex)

Liver_Disease_Mortality_Burden_Race <- Liver_Disease_Mortality_Burden |>
  filter(Characteristic %in% c(Races))

print(Liver_Disease_Mortality_Burden_Race)
```

```{r, warning=FALSE}
#| label: Cleaning kidney disease mortality burden data and creating by race/sex data frames

# Renaming columns
Kidney_Disease_Mortality_Burden_Columns <- as.character(Kidney_Disease_Mortality_Burden[3, ])
Kidney_Disease_Mortality_Burden_Columns[is.na(Kidney_Disease_Mortality_Burden_Columns)] <- "NA"
Kidney_Disease_Mortality_Burden_Columns <- make.unique(Kidney_Disease_Mortality_Burden_Columns)
  # make.unique appends a sequence of numbers to make the column names unique
colnames(Kidney_Disease_Mortality_Burden) <- Kidney_Disease_Mortality_Burden_Columns

Kidney_Disease_Mortality_Burden <- Kidney_Disease_Mortality_Burden |>
  rename(Characteristic = "Sex and cause of death\\1") |>
  select(Characteristic,`1997\\3`:`2019\\4`)

# Removing irrelevant data
Kidney_Disease_Mortality_Burden <- Kidney_Disease_Mortality_Burden[c(21, 42, 63, 104, 144, 224, 284), ]

# Updating characteristics for "Chronic liver disease and cirrhosis"
Kidney_Disease_Mortality_Burden[1, "Characteristic"] <- "All"
Kidney_Disease_Mortality_Burden[2, "Characteristic"] <- "Male"
Kidney_Disease_Mortality_Burden[3, "Characteristic"] <- "Female"
Kidney_Disease_Mortality_Burden[4, "Characteristic"] <- "White only"
Kidney_Disease_Mortality_Burden[5, "Characteristic"] <- "Black or African American only"
Kidney_Disease_Mortality_Burden[6, "Characteristic"] <- "Asian or Pacific Islander only"
Kidney_Disease_Mortality_Burden[7, "Characteristic"] <- "Hispanic or Latino"

# Cleaning year names
colnames(Kidney_Disease_Mortality_Burden) <- gsub("\\\\[0-9]$", "", colnames(Kidney_Disease_Mortality_Burden))

Kidney_Disease_Mortality_Burden <- Kidney_Disease_Mortality_Burden |>
  mutate(across(`1997`:`2019`, ~ as.numeric(.x) / 100000))

print(Kidney_Disease_Mortality_Burden)

Kidney_Disease_Mortality_Burden_Sex <- Kidney_Disease_Mortality_Burden |>
  filter(Characteristic %in% c(Sexes))

print(Kidney_Disease_Mortality_Burden_Sex)

Kidney_Disease_Mortality_Burden_Race <- Kidney_Disease_Mortality_Burden |>
  filter(Characteristic %in% c(Races))

print(Kidney_Disease_Mortality_Burden_Race)
```

## Donor Living Status (only applicable to livers and kidneys)
```{r}
#| label: Creating data frame for proportions of living v deceased organ donations

Categories <- c("1 - Donor Type", "0 - Total")
Donor_Living_Status <- c("Deceased Donor", "Living Donor")

# Filtering for just data on living status
Donor_Living_Status <- HRSA_Donor_Demographics |>
  filter(Category %in% Categories) |>
  select(`Donor Type Filter`, `Donor Count`) |>
  rename(Status = `Donor Type Filter`,
         Count = `Donor Count`) |>
  mutate(Count = as.numeric(Count))

# Saving total value
All <- Donor_Living_Status$Count[Donor_Living_Status$Status == "Grand Total"]

# Add proportion column
Donor_Living_Status <- Donor_Living_Status |>
  mutate(Proportion = (Count / All))   

print(Donor_Living_Status)
```

## Recipient Demographic Data
```{r, warning = FALSE}
#| label: Heart recipient demographic data

Heart_Recipients <- HRSA_Recipient_Demographics |>
  filter(`Organ Transplanted` == "Heart") |>
  select(`Transplant Year`:`Distinct Recipients with Other Insurance Payment Sources`)

# Generating proportions from counts
Heart_Recipients <- Heart_Recipients |>
  mutate(across(`Total Distinct Recipients`:`Distinct Recipients with Other Insurance Payment Sources`, 
                ~ as.numeric(.x))) |>
  mutate(across(`Male Distinct Recipients`:`Distinct Recipients with Other Insurance Payment Sources`, 
                ~ .x / `Total Distinct Recipients`,
                .names = "Prop_{.col}"))

# Renaming Prop_ variables
Heart_Recipients <- Heart_Recipients |>
  rename_with(
    .fn = ~ paste0(gsub(" Distinct Recipients", "", .x)),
    .cols = `Prop_Male Distinct Recipients`:`Prop_Distinct Recipients with Other Insurance Payment Sources`
  ) |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_Male:`Prop_Distinct Recipients with Other Insurance Payment Sources`)

print(Heart_Recipients)
```


```{r, warning=FALSE}
#| label: Adding proportions to liver and kidney recipient demographic data

Organs <- c("Liver", "Kidney")

Deceased_Organ_Recipients <- HRSA_Recipient_Demographics |>
  filter(`Organ Transplanted` %in% Organs,
         `Donor Type` == "Deceased Donor")

Living_Organ_Recipients <- HRSA_Recipient_Demographics |>
  filter(`Organ Transplanted` %in% Organs,
         `Donor Type` == "Living Donor")

Deceased_v_Living_Organ_Recipients <- rbind(Deceased_Organ_Recipients, Living_Organ_Recipients)

Deceased_v_Living_Organ_Recipients <- Deceased_v_Living_Organ_Recipients |>
  select(`Transplant Year`:`Distinct Recipients with Other Insurance Payment Sources`)

# Generating proportions from counts
Deceased_v_Living_Organ_Recipients <- Deceased_v_Living_Organ_Recipients |>
  mutate(across(`Total Distinct Recipients`:`Distinct Recipients with Other Insurance Payment Sources`, 
                ~ as.numeric(.x))) |>
  mutate(across(`Male Distinct Recipients`:`Distinct Recipients with Other Insurance Payment Sources`, 
                ~ .x / `Total Distinct Recipients`,
                .names = "Prop_{.col}"))

# Renaming Prop_ variables
Deceased_v_Living_Organ_Recipients <- Deceased_v_Living_Organ_Recipients |>
  rename_with(
    .fn = ~ paste0(gsub(" Distinct Recipients", "", .x)),
    .cols = `Prop_Male Distinct Recipients`:`Prop_Distinct Recipients with Other Insurance Payment Sources`
  )
```



```{r, warning=FALSE}
#| label: Liver recipient demographic data

Deceased_v_Living_Liver_Recipients <- Deceased_v_Living_Organ_Recipients |>
  filter(`Organ Transplanted` == "Liver") |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_Male:`Prop_Distinct Recipients with Other Insurance Payment Sources`)

print(Deceased_v_Living_Liver_Recipients)
```

```{r, warning=FALSE}
#| label: Kidney ONLY recipient demographic data

# Note: Pancreases are almost exclusively harvested from deceased donors, so excluding them

Deceased_v_Living_Kidney_Recipients <- Deceased_v_Living_Organ_Recipients |>
  filter(`Organ Transplanted` != "Liver") |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_Male:`Prop_Distinct Recipients with Other Insurance Payment Sources`)

print(Deceased_v_Living_Kidney_Recipients)
```

# Exploratory Data Analysis

## Disease Mortality Burden

### Cardiovascular Disease Mortality Burden
```{r, warning=FALSE}
#| label: Cardiovascular disease mortality by sex

print(Cardiovascular_Disease_Mortality_Burden_Sex)

Cardiovascular_Disease_Mortality_Burden_Sex_Plot_Data <- Cardiovascular_Disease_Mortality_Burden_Sex |>
  select(Sex, `2010`:`2019`) |>
  pivot_longer(cols = `2010`:`2019`,
               names_to = "Year",
               values_to = "Prevalence") |>
  mutate(
    Prevalence = as.numeric(Prevalence),
    Year = Year
  )

print(Cardiovascular_Disease_Mortality_Burden_Sex_Plot_Data)

ggplot(Cardiovascular_Disease_Mortality_Burden_Sex_Plot_Data, aes(x = Year, y = Prevalence, fill = Sex)) +
  geom_col(position = "dodge") +
scale_y_continuous(
  limits = c(0, 0.01),
  breaks = seq(0, 0.01, by = 0.002),
  labels = scales::percent_format(accuracy = 0.1)
) +
  labs(title = "Cardiovascular Disease Mortality by Sex",
       y = "Prevalence",
       x = "Year") +
  theme_minimal()
```

```{r, warning=FALSE}
#| label: Cardiovascular disease mortality by race

print(Cardiovascular_Disease_Mortality_Burden_Race)

Cardiovascular_Disease_Mortality_Burden_Sex_Plot_Data <- Cardiovascular_Disease_Mortality_Burden_Race |>
  select(Race, `2010`:`2019`) |>
  pivot_longer(cols = `2010`:`2019`,
               names_to = "Year",
               values_to = "Prevalence") |>
  mutate(
    Prevalence = as.numeric(Prevalence),
    Year = Year
  )

print(Cardiovascular_Disease_Mortality_Burden_Sex_Plot_Data)

ggplot(Cardiovascular_Disease_Mortality_Burden_Sex_Plot_Data, aes(x = Year, y = Prevalence, fill = Race)) +
  geom_col(position = "dodge") +
scale_y_continuous(
  limits = c(0, 0.01),
  breaks = seq(0, 0.01, by = 0.002),
  labels = scales::percent_format(accuracy = 0.1)
) +
  labs(title = "Cardiovascular Disease Mortality by Race",
       y = "Prevalence",
       x = "Year") +
  theme_minimal()
```

### Liver Disease Mortality Burden

```{r, warning=FALSE}
#| label: Liver disease mortality by sex

print(Liver_Disease_Mortality_Burden_Sex)

Liver_Disease_Mortality_Burden_Sex_Plot_Data <- Liver_Disease_Mortality_Burden_Sex |>
  select(Characteristic, `2010`:`2019`) |>
  rename(Sex = Characteristic) |>
  pivot_longer(cols = `2010`:`2019`,
               names_to = "Year",
               values_to = "Prevalence") |>
  mutate(
    Prevalence = as.numeric(Prevalence),
    Year = Year
  )

print(Liver_Disease_Mortality_Burden_Sex_Plot_Data)

ggplot(Liver_Disease_Mortality_Burden_Sex_Plot_Data, aes(x = Year, y = Prevalence, fill = Sex)) +
  geom_col(position = "dodge") +
scale_y_continuous(
  limits = c(0, 0.001),
  breaks = seq(0, 0.001, by = 0.0002),
  labels = scales::percent_format(accuracy = 0.01)
) +
  labs(title = "Liver Disease Mortality by Sex",
       y = "Prevalence",
       x = "Year") +
  theme_minimal()
```

```{r, warning=FALSE}
#| label: Liver disease mortality by race

print(Liver_Disease_Mortality_Burden_Race)

Liver_Disease_Mortality_Burden_Race_Plot_Data <- Liver_Disease_Mortality_Burden_Race |>
  select(Characteristic, `2010`:`2019`) |>
  rename(Race = Characteristic) |>
  pivot_longer(cols = `2010`:`2019`,
               names_to = "Year",
               values_to = "Prevalence") |>
  mutate(
    Prevalence = as.numeric(Prevalence),
    Year = Year
  )

print(Liver_Disease_Mortality_Burden_Race_Plot_Data)

ggplot(Liver_Disease_Mortality_Burden_Race_Plot_Data, aes(x = Year, y = Prevalence, fill = Race)) +
  geom_col(position = "dodge") +
scale_y_continuous(
  limits = c(0, 0.001),
  breaks = seq(0, 0.001, by = 0.0002),
  labels = scales::percent_format(accuracy = 0.01)
) +
  labs(title = "Liver Disease Mortality by Race",
       y = "Prevalence",
       x = "Year") +
  theme_minimal()
```

### Kidney Disease Mortality Burden

```{r, warning=FALSE}
#| label: Kidney disease mortality by sex

print(Kidney_Disease_Mortality_Burden_Sex)

Kidney_Disease_Mortality_Burden_Sex_Plot_Data <- Kidney_Disease_Mortality_Burden_Sex |>
  select(Characteristic, `2010`:`2019`) |>
  rename(Sex = Characteristic) |>
  pivot_longer(cols = `2010`:`2019`,
               names_to = "Year",
               values_to = "Prevalence") |>
  mutate(
    Prevalence = as.numeric(Prevalence),
    Year = Year
  )

print(Kidney_Disease_Mortality_Burden_Sex_Plot_Data)

ggplot(Kidney_Disease_Mortality_Burden_Sex_Plot_Data, aes(x = Year, y = Prevalence, fill = Sex)) +
  geom_col(position = "dodge") +
scale_y_continuous(
  limits = c(0, 0.001),
  breaks = seq(0, 0.001, by = 0.0002),
  labels = scales::percent_format(accuracy = 0.01)
) +
  labs(title = "Kidney Disease Mortality by Sex",
       y = "Prevalence",
       x = "Year") +
  theme_minimal()
```

```{r, warning=FALSE}
#| label: Kidney disease mortality by race

print(Kidney_Disease_Mortality_Burden_Race)

Kidney_Disease_Mortality_Burden_Race_Plot_Data <- Kidney_Disease_Mortality_Burden_Race |>
  select(Characteristic, `2010`:`2019`) |>
  rename(Race = Characteristic) |>
  pivot_longer(cols = `2010`:`2019`,
               names_to = "Year",
               values_to = "Prevalence") |>
  mutate(
    Prevalence = as.numeric(Prevalence),
    Year = Year
  )

print(Kidney_Disease_Mortality_Burden_Race_Plot_Data)

ggplot(Kidney_Disease_Mortality_Burden_Race_Plot_Data, aes(x = Year, y = Prevalence, fill = Race)) +
  geom_col(position = "dodge") +
scale_y_continuous(
  limits = c(0, 0.001),
  breaks = seq(0, 0.001, by = 0.0002),
  labels = scales::percent_format(accuracy = 0.01)
) +
  labs(title = "Kidney Disease Mortality by Race",
       y = "Prevalence",
       x = "Year") +
  theme_minimal()
```

# Organ Allocation

## Heart Allocation
```{r}
#| label: Heart recipients by sex

Heart_Plot_Sex_Data <- Heart_Recipients |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_Male, Prop_Female) |>
  pivot_longer(cols = starts_with("Prop_"),
               names_to = "Sex",
               values_to = "Proportion") |>
  mutate(Sex = str_remove(Sex, "Prop_"))

ggplot(Heart_Plot_Sex_Data, aes(x = `Transplant Year`, y = Proportion, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.1)) +
  labs(title = "Proportion of Heart Transplant Recipients by Sex",
       y = "Proportion",
       x = "Year") +
  theme_minimal()
```

```{r}
#| label: Heart recipients by race

Heart_Plot_Race_Data <- Heart_Recipients |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_White:Prop_Multiracial) |>
  pivot_longer(cols = starts_with("Prop_"),
               names_to = "Race",
               values_to = "Proportion") |>
  mutate(Race = str_remove(Race, "Prop_")) |>
  filter(!Race %in% c("Multiracial", "Native Hawaiian", "American Indian"))

ggplot(Heart_Plot_Race_Data, aes(x = `Transplant Year`, y = Proportion, fill = Race)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.1)
                     ) +
  labs(title = "Proportion of Heart Transplant Recipients by Race",
       y = "Proportion",
       x = "Year") +
  theme_minimal()
```


## Liver Allocation
```{r}
#| label: Livers recipients by sex

Liver_Plot_Sex_Data <- Deceased_v_Living_Liver_Recipients |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_Male, Prop_Female) |>
  pivot_longer(cols = starts_with("Prop_"),
               names_to = "Sex",
               values_to = "Proportion") |>
  mutate(Sex = str_remove(Sex, "Prop_"))


ggplot(Liver_Plot_Sex_Data, aes(x = `Transplant Year`, y = Proportion, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ `Donor Type`) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.1)
                     ) +
  labs(title = "Proportion of Liver Transplant Recipients by Sex",
       y = "Proportion",
       x = "Year") +
  theme_minimal()

# It appear that males get deceased donor kidneys more often than females
```

```{r, warning = FALSE}
#| label: Liver recipients by race

Liver_Plot_Race_Data <- Deceased_v_Living_Liver_Recipients |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_White:Prop_Multiracial) |>
  pivot_longer(cols = starts_with("Prop_"),
               names_to = "Race",
               values_to = "Proportion") |>
  mutate(Race = str_remove(Race, "Prop_")) |>
  filter(!Race %in% c("Multiracial", "Native Hawaiian", "American Indian"))


ggplot(Liver_Plot_Race_Data, aes(x = `Transplant Year`, y = Proportion, fill = Race)) +
  geom_col(position = "dodge") +
  facet_wrap(~ `Donor Type`) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.1)
                     ) +
  labs(title = "Proportion of Liver Transplant Recipients by Race",
       y = "Proportion",
       x = "Year") +
  theme_minimal()
```


## Kidney Allocation
```{r}
#| label: Kidney recipients by sex

Kidney_Plot_Sex_Data <- Deceased_v_Living_Kidney_Recipients |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_Male, Prop_Female) |>
  pivot_longer(cols = starts_with("Prop_"),
               names_to = "Sex",
               values_to = "Proportion") |>
  mutate(Sex = str_remove(Sex, "Prop_"))


ggplot(Kidney_Plot_Sex_Data, aes(x = `Transplant Year`, y = Proportion, fill = Sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ `Donor Type`) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.1)
                     ) +
  labs(title = "Proportion of Kidney Transplant Recipients by Sex",
       y = "Proportion",
       x = "Year") +
  theme_minimal()

# It appear that males get deceased donor kidneys more often than females
```

```{r, warning = FALSE}
#| label: Kidney recipients by race

Kidney_Plot_Race_Data <- Deceased_v_Living_Kidney_Recipients |>
  select(`Transplant Year`:`Organ Transplanted`, Prop_White:Prop_Multiracial) |>
  pivot_longer(cols = starts_with("Prop_"),
               names_to = "Race",
               values_to = "Proportion") |>
  mutate(Race = str_remove(Race, "Prop_")) |>
  filter(!Race %in% c("Multiracial", "Native Hawaiian", "American Indian"))

ggplot(Kidney_Plot_Race_Data, aes(x = `Transplant Year`, y = Proportion, fill = Race)) +
  geom_col(position = "dodge") +
  facet_wrap(~ `Donor Type`) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), breaks = seq(0, 1, by = 0.1)
                     ) +
  labs(title = "Proportion of Kidney Transplant Recipients by Race",
       y = "Proportion",
       x = "Year") +
  theme_minimal()
```


### Cardiovascular
```{r}
#| label: Testing for a difference in cardiovascular disease mortality by race

print(Cardiovascular_Disease_Mortality_Burden_Race)

# Pivoting to long
Cardiovascular_Disease_Mortality_Burden_Race_Test <- Cardiovascular_Disease_Mortality_Burden_Race |>
  pivot_longer(
    cols = `1997`:`2019`,
    names_to = "Year",
    values_to = "Prevalence"
  ) 

print(Cardiovascular_Disease_Mortality_Burden_Race_Test)

# Hypotheses:
# H0: Mean cardiovascular disease mortality is the same across racial groups
# H1: Mean cardiovascular disease mortality differs between at least one pair of racial groups

ANOVA_Cardiovascular_Disease_Mortality_Burden_Race <- aov(as.numeric(Prevalence) ~ Race, data = Cardiovascular_Disease_Mortality_Burden_Race_Test)

summary(ANOVA_Cardiovascular_Disease_Mortality_Burden_Race)

# Conclusion: There are significant differences in cardiovascular disease mortality between at least one pair of racial groups

Cardiovascular_Disease_Mortality_Burden_Race_Test %>%
  group_by(Race) %>%
  summarise(
    count = n(),
    mean = base::mean(as.numeric(Prevalence), na.rm=TRUE),
    sd = stats::sd(as.numeric(Prevalence), na.rm=TRUE),
    min = base::min(as.numeric(Prevalence), na.rm=TRUE),
    max = base::max(as.numeric(Prevalence), na.rm=TRUE)
  )
```

### Liver
```{r}
#| label: Testing for a difference in liver disease mortality by race

print(Liver_Disease_Mortality_Burden_Race)

# Pivoting to long
Liver_Disease_Mortality_Burden_Race_Test <- Liver_Disease_Mortality_Burden_Race |>
  pivot_longer(
    cols = `1997`:`2019`,
    names_to = "Year",
    values_to = "Prevalence"
  ) |>
  rename(Race = Characteristic)

print(Liver_Disease_Mortality_Burden_Race_Test)

# Summary statistics by Race
Liver_Disease_Mortality_Burden_Race_Test |>
  group_by(Race) |>
  summarise(
    count = n(),
    mean = base::mean(Prevalence, na.rm = TRUE),
    sd = stats::sd(Prevalence, na.rm = TRUE),
    min = base::min(Prevalence, na.rm = TRUE),
    max = base::max(Prevalence, na.rm = TRUE)
  ) |> print()

# Hypotheses:
# H0: Mean liver disease mortality is the same across racial groups
# H1: Mean liver disease mortality differs between at least one pair of racial groups

ANOVA_Liver_Disease_Mortality_Burden_Race <- aov(as.numeric(Prevalence) ~ Race, data = Liver_Disease_Mortality_Burden_Race_Test)

summary(ANOVA_Liver_Disease_Mortality_Burden_Race)

# Conclusion: There are significant differences in liver disease mortality between among at least one pair of racial groups

```

### Kidney
```{r}
#| label: Testing for a difference in kidney disease mortality by race

print(Kidney_Disease_Mortality_Burden_Race)

# Pivoting to long
Kidney_Disease_Mortality_Burden_Race_Test <- Kidney_Disease_Mortality_Burden_Race |>
  pivot_longer(
    cols = `1997`:`2019`,
    names_to = "Year",
    values_to = "Prevalence"
  ) |>
  rename(Race = Characteristic)

print(Kidney_Disease_Mortality_Burden_Race_Test)

# Summary statistics by Race
Kidney_Disease_Mortality_Burden_Race_Test |>
  group_by(Race) |>
  summarise(
    count = n(),
    mean = base::mean(Prevalence, na.rm = TRUE),
    sd = stats::sd(Prevalence, na.rm = TRUE),
    min = base::min(Prevalence, na.rm = TRUE),
    max = base::max(Prevalence, na.rm = TRUE)
  ) |> print()

# Hypotheses:
# H0: Mean kidney disease mortality is the same across racial groups
# H1: Mean kidney disease mortality differs between at least one pair of racial groups

ANOVA_Kidney_Disease_Mortality_Burden_Race_Test <- aov(as.numeric(Prevalence) ~ Race, data = Kidney_Disease_Mortality_Burden_Race_Test)

summary(ANOVA_Kidney_Disease_Mortality_Burden_Race_Test)

# Conclusion: There are significant differences in kidney disease mortality between among at least one pair of racial groups
```

